---
title: "Homework 4"
author: "Yuki Joyama"
date: "2024-02-29"
output: 
  pdf_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)

library(tidyverse)
library(MASS)
library(nnet)
```

```{r}
# data prep
df_house = data.frame(
  contact = c(rep(c("low", "high"), times = c(3, 3))),
  type = c(rep(c("tower", "apartment", "house"), length.out = 3)), 
  sat.low = c(65, 130, 67, 34, 141, 130),
  sat.medium = c(54, 76, 48, 47, 116, 105),
  sat.high = c(100, 111, 62, 100, 191, 104)
)
```

## 1
```{r}
# calculate row-wise percentages
df_house$sat.low_per <- (df_house$sat.low / rowSums(df_house[, c("sat.low", "sat.medium", "sat.high")])) * 100
df_house$sat.medium_per <- (df_house$sat.medium / rowSums(df_house[, c("sat.low", "sat.medium", "sat.high")])) * 100
df_house$sat.high_per <- (df_house$sat.high / rowSums(df_house[, c("sat.low", "sat.medium", "sat.high")])) * 100

# table of percentages
df_house[-(3:5)]

# plot
df_house |> 
  dplyr::select(contact, type, sat.low_per, sat.medium_per, sat.high_per) |> 
  pivot_longer(cols = starts_with("sat."), 
               names_to = "sat", 
               values_to = "proportion") |> 
  mutate(sat = str_remove(sat, "sat\\.") |>  str_remove("_per")) |> 
# plot
  ggplot(aes(x = type, y = proportion, group = sat, color = sat)) +
  geom_line() +
  facet_grid(~contact) +
  theme_bw()
```

Percentages of responses in each category by contact with other residents and type of housing is summarized in the above table and plots. Top panel is the group that answered "low" for contact and bottom is the group that answered "high". We can see that tower residents are likely to have high satisfaction compared to other types of housing, and high contact tends to have high satisfaction expect for those who live in a house. 

## 2
```{r results=F}
# fit a nominal logistic regression model
house.mult <- multinom(cbind(sat.low, sat.medium, sat.high) ~ factor(contact) + factor(type), data = df_house)
```

```{r}
summary(house.mult)

# obtain odds ratio for each coefficient
exp(coef(house.mult))
```

The nominal logistic regression model is as follows:  
$log(\frac{\pi_j}{\pi_1})=\beta_{0j}+\beta_{1j}x_1+\beta_{2j}x_2+\beta_{3j}x_3$, $j = 2,3$   

where  
$\pi_1:$ the probability of low satisfaction   
$\pi_2:$ the probability of medium satisfaction  
$\pi_3:$ the probability of high satisfaction   

$$
x_1=
\begin{cases}
1&\text{for low contact}\\
0&\text{for high contact}\\
\end{cases}
$$
$$
x_2=
\begin{cases}
1&\text{for house}\\
0&\text{for apartment}\\
\end{cases}
$$
$$
x_3=
\begin{cases}
1&\text{for tower}\\
0&\text{for apartment}\\
\end{cases}
$$
We observe that when the level of contact with other residents is low, the odds of experiencing medium and high satisfaction decrease compared to low satisfaction, while controlling for other variables. Similarly, for individuals residing in houses compared to apartments, the odds of experiencing medium satisfaction are higher relative to low satisfaction, whereas the odds of high satisfaction are lower holding other variables unchanged. Among those living in towers, the odds of experiencing both medium and high satisfaction are higher than those for low satisfaction, all else being equal.  
For example, the odds ratio of falling into high satisfaction category (vs low satisfaction) for tower residents with high contact with other residents is `r round(exp(coef(house.mult))[2, 4], 2)`  

The 95% confidence intervals for each odds ratio is as follows:
```{r}
# 95%CI for odds ratio
exp(confint(house.mult))
```


```{r}
# goodness of fit
pihat = predict(house.mult,type = 'probs') 
pihat
m = rowSums(df_house[, 3:5])

# pearson residuals 
res.pearson = (df_house[, 3:5] - pihat*m) / sqrt(pihat*m) 
res.pearson

# Generalized Pearson Chisq Stat
G.stat = sum(res.pearson^2) 
G.stat
pval = 1 - pchisq(G.stat, df = (6 - 4)*(3 - 1))
pval 

# deviance
D.stat = sum(2*df_house[, 3:5]*log(df_house[, 3:5] / (m*pihat)))
D.stat
```

Generalized Pearson $\chi^2$ statistic shows p-value of `r round(pval, 2)`, indicating that the model has a good fit.

```{r}
# interaction
house.mult_int <- multinom(cbind(sat.low, sat.medium, sat.high) ~ factor(contact) + factor(type) + factor(contact)*factor(type), data = df_house)

summary(house.mult_int)

# obtain odds ratio for each coefficient
exp(coef(house.mult_int))

# 95%CI for odds ratio
exp(confint(house.mult_int))
```

From the output, we can see that there is a statistically significant interaction between low contact and house type for odds ratio of high satisfaction vs low satisfaction.

## 3
Now, I will treat the satisfaction categories as ordinal variable and fit a proportional odds model. 
```{r}
# data prep
df_house2 <- df_house |> 
  dplyr::select(contact, type, sat.low, sat.medium, sat.high) |> 
  pivot_longer(cols = starts_with("sat."), 
               names_to = "sat", 
               values_to = "frequency") |> 
  mutate(
    sat = str_remove(sat, "sat\\."), 
    sat = factor(sat, levels = c("low", "medium", "high"))
  )

# fit an ordinal logistic regression model
house.mult2 <- polr(sat ~ contact + type, data = df_house2, weights = frequency)
summary(house.mult2)
exp(-coef(house.mult2))
```

Let $Y$ be an ordinal outcome with $J$ categories. $P(Y\leq j)$ is the cumulative probability of $Y$ less than or equal to a specific category $j=1,...,J-1$  
In polr, the ordinal logistic regression model is parameterized as  
$logit(P(Y\leq j))=log\frac{P(Y\leq j)}{P(Y>j)}=\beta_{j0}-\eta_1x_1-\eta_2x_2-\eta_3x_3$  
where $\eta_i=-\beta_i$  

The output shows that the odds ratio of falling in lower category is `r round(exp(-coef(house.mult2))[1], 2)` in people with low contact holding other variables constant. And the odds ratio of falling in lower category is `r round(exp(-coef(house.mult2))[3], 2)` in people living in tower all else being equal.

## 4
```{r}
# Pearson residuals 
pihat = predict(house.mult2, df_house2, type = 'p')
m = rowSums(cbind(df_house[, 3:5]))
res.pearson = (df_house[, 3:5] - pihat*m) / sqrt(pihat*m)
G = sum(res.pearson^2)
G

numsamp = (3 - 1)*6 # degree of freedom for grouped data
numparam = 2 + 3 # total num of param
pval = 1 - pchisq(G, df = numsamp - numparam)
pval # fits well
```









