---
title: "Homework 3"
author: "Yuki Joyama"
date: "2024-02-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)

library(tidyverse)
```

## 1
```{r}
# data prep
df_ec = data.frame(
  age = c(25, 25, 35, 35, 45, 45, 55, 55, 65, 65, 75, 75),
  alcohol = c(rep(c("less", "more"), length.out = 12)),
  case = c(0, 1, 5, 4, 21, 25, 24, 42, 36, 19, 8, 5),
  control = c(106, 9, 164, 26, 138, 29, 139, 27, 88, 18, 31, 0)
) 

resp = cbind(df_ec$case, df_ec$control) 
```

We coded daily alcohol consumption 0-79g as "less", and 80+g as "more".  

### a 
```{r}
# fit a prospective logit model
logit.prosp = glm(resp ~ df_ec$alcohol + df_ec$age, family = binomial(link = 'logit'))

summary(logit.prosp) 

exp(coef(logit.prosp)) # odds ratio estimates
```

In this model, we treat disease status (case vs control) as response and exposures (daily alcohol consumption and age) as predictors.  

$logit(\pi_i)=\beta_0+\beta_1x_1+\beta_2x_2$  
$x_1:$ the indicator of heavier daily alcohol consumption (80+g)  
$x_2:$ the indicator of age (as a continuous variable)  

* The odds ratio of having esophageal cancer is `r round(exp(coef(logit.prosp))[2], 3)` in heavier daily alcohol consumer (80+g) compared to lighter consumer (0-79g) holding other covariates constant   
* As age increases by one year, the odds ratio of having esophageal cancer increases by `r round(exp(coef(logit.prosp))[3], 3)` holding other covariates constant   

### b

