---
title: "Homework 7"
author: "Yuki Joyama"
date: "2024-04-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)

library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)

# setup plot theme
theme_set(
  theme_bw() +
    theme(legend.position = "top")
  )
```

```{r data}
# data prep
df = read_csv("HW7-politeness_data.csv")
```

## (a) Exploratory Analysis

```{r boxplot}
# gender/attitude and pitch
df |> 
  ggplot(aes(x = attitude, y = frequency, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~gender)

gender <- df$gender
attitude <- df$attitude
subject <- df$subject
frequency <- df$frequency
scenario <- df$scenario
```

The boxplots illustrates the relation between gender/attitude and pitch. We can see that Female tend to have higher frequency than Male, and lower frequency is more likely to be considered as formal (pol). 

## (b) LMM with random intercepts
I will fit a mixed effects model with random intercepts for different subjects with gender and attitude being the fixed effects.  
```{r}
# LMM with random intercept
# gender, attitude fixed
LMM1 <- lme(frequency ~ gender + attitude, random = ~1 | subject, method = "REML")

summary(LMM1)
```

```{r}
VarCorr(LMM1)
```

Given the output, 
$$
cov(Y_i) = \begin{pmatrix}
1445.90 & 598.20 & ... & 598.20 \\
598.20 & 1445.90 & ... & 598.20\\
...\\
598.20 & 598.20 & ... & 1445.90
\end{pmatrix}
$$

The covariance matrix for the estimates of fixed effects are as follows:
```{r}
vcov(LMM1)
```

The best linear unbiased predictions (BLUPs) for subject-specific intercepts:
```{r}
random.effects(LMM1)
```

Residuals: 
```{r}
resid(LMM1)
hist(resid(LMM1))
```

# (c) LMM with random intercepts and interaction
I will fit a mixed effects model with random intercepts for different subjects with gender, attitude and their interaction being the fixed effects.  
```{r}
# LMM with random intercept
# gender, attitude, gender*attitude fixed
LMM2 <- lme(frequency ~ gender + attitude + gender*attitude, random = ~1 | subject, method = "REML")

summary(LMM2)
```
The output indicates that the interaction term does not have a significant influence on the response variable.

Now, I will refit LMM1 and LMM2 using ML method for the likelihood ratio test.
```{r}
LMM1.1 <- lme(frequency ~ gender + attitude, random = ~1 | subject, method = "ML")
LMM2.1 <- lme(frequency ~ gender + attitude + gender*attitude, random = ~1 | subject, method = "ML")

# LRT of interaction 
anova(LMM1.1, LMM2.1)
```
Given the result (p-value > 0.05), we fail to reject the null hypothesis. Therefore, we conclude that including the interaction term is not significantly associated with pitch. 

# (d) LMM with random intercept for both subject and scenarios
I will fit a mixed effects model with random intercepts for different subjects and scenarios with gender and attitude being the fixed effects. 
```{r}
LMM3 <- lmer(frequency ~ gender + attitude + (1 | subject) + (1 | scenario))

summary(LMM3) 
```

```{r}
print(VarCorr(LMM3), comp = "Variance")
```

Given the output, 
$$
cov(Y_i) = \begin{pmatrix}
1475.47 & 837.69 & ... & 837.69 \\
837.69 & 1475.47 & ... & 837.69\\
...\\
837.69 & 837.69& ... & 1475.47
\end{pmatrix}
$$
The coefficient for the fixed effect `attitudepol` is -20.00. This means that the pitch is lower in polite speech than in informal speech, by about 20 Hz holding other variable constant. 



