---
title: "Homework 5"
author: "Yuki Joyama"
date: "2024-03-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)

library(tidyverse)
```

```{r data}
# data prep 
df_crab = read.table("/Users/yukijoyama/Library/CloudStorage/GoogleDrive-jikeyu1995@gmail.com/My Drive/version_control/MS-biostats/Biostatistical Methods 2/Biostatistical Methods 2/p8131_hw5_files/HW5-crab.txt", header = TRUE)
df_para = read.table("/Users/yukijoyama/Library/CloudStorage/GoogleDrive-jikeyu1995@gmail.com/My Drive/version_control/MS-biostats/Biostatistical Methods 2/Biostatistical Methods 2/p8131_hw5_files/HW5-parasite.txt", header = TRUE)
```

# 1
## (a)
I will fit a Poisson model (M1) with log link with carapace width (W) as the single predictor.

```{r}
# M1: Poisson model with log link
m1_fit <- glm(Sa ~ W, family = poisson(link = "log"), data = df_crab)
summary(m1_fit)
exp(m1_fit$coefficients)
```

The expected number of satellites (Sa) given carapace width (W) = 0 is `r round(exp(m1_fit$coefficients)[1], 3)`. With every unit increase in W, the expected number of Sa has multiplicative effect of `r round(exp(m1_fit$coefficients)[2], 3)` on $\mu=E(Y)$.

```{r}
# Goodness of fit
G = sum(residuals(m1_fit, type = "pearson") ^ 2) 
G 
1 - pchisq(G, 2)
```

Thus, we reject the null. The model does not have a good fit. 

## (b)
Now, I will fit a Poisson model (M2) with log link with carapace width (W) and weight (Wt) as predictors.

```{r}
# M2: W and Wt as predictors
m2_fit <- glm(Sa ~ W + Wt, family = poisson(link = "log"), data = df_crab)
summary(m2_fit)
exp(m2_fit$coefficients)
```

```{r}
# compare with M1
anova(m1_fit, m2_fit)
```

## (c)
```{r}
# over-dispersion in M2

```

```{r}
# adjust for over-dispersion
```


# 2
## a

## b

## c

