---
title: "Homework 5"
author: "Yuki Joyama"
date: "2024-03-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)

library(tidyverse)
library(pscl) # zip model
library(performance)
```

```{r data}
# data prep 
df_crab = read.table("/Users/yukijoyama/Library/CloudStorage/GoogleDrive-jikeyu1995@gmail.com/My Drive/version_control/MS-biostats/Biostatistical Methods 2/Biostatistical Methods 2/p8131_hw5_files/HW5-crab.txt", header = TRUE)
df_para = read.table("/Users/yukijoyama/Library/CloudStorage/GoogleDrive-jikeyu1995@gmail.com/My Drive/version_control/MS-biostats/Biostatistical Methods 2/Biostatistical Methods 2/p8131_hw5_files/HW5-parasite.txt", header = TRUE)
```

# 1
## (a)
I will fit a Poisson model (M1) with log link with carapace width (W) as the single predictor.

```{r}
# M1: Poisson model with log link
m1_fit <- glm(Sa ~ W, family = poisson(link = "log"), data = df_crab)
summary(m1_fit)
exp(m1_fit$coefficients)
```

The expected number of satellites (Sa) given carapace width (W) = 0 is `r round(exp(m1_fit$coefficients)[1], 3)`. With every unit increase in W, the expected number of Sa has multiplicative effect of `r round(exp(m1_fit$coefficients)[2], 3)` on $\mu=E(Y)$.

```{r}
# Goodness of fit
G = sum(residuals(m1_fit, type = "pearson") ^ 2) 
G 
1 - pchisq(G, m1_fit$df.residual)
```

Thus, we reject the null, which indicates that the model does not have a good fit. 

## (b)
Now, I will fit a Poisson model (M2) with log link with carapace width (W) and weight (Wt) as predictors.

```{r}
# M2: W and Wt as predictors
m2_fit <- glm(Sa ~ W + Wt, family = poisson(link = "log"), data = df_crab)
summary(m2_fit)
exp(m2_fit$coefficients)
```

The expected number of Sa given W = 0 and Wt = 0 is `r round(exp(m2_fit$coefficients)[1], 3)`. All else being equal, with every unit increase in W, the expected number of Sa has multiplicative effect of `r round(exp(m2_fit$coefficients)[2], 3)` on $\mu=E(Y)$. Similarly, holding other variables constant, with every unit increase in Wt, the expected number of Sa has multiplicative effect of `r round(exp(m2_fit$coefficients)[3], 3)` on $\mu=E(Y)$.   

M1 is nested within M2, so I will perform ANOVA to compare the two models. 
```{r}
# compare with M1
anova(m1_fit, m2_fit)
```

The residual deviance appears to be reduced in M2 compared to M1, indicating that M2 has a better fit by adding Wt as a predictor. We need to note that the coefficient in W in M2 is not statistically significant unlike M1.

## (c)
```{r}
# over-dispersion in M2
# calculate dispersion parameter
G.stat = sum(residuals(m2_fit, type = 'pearson', data = df_crab) ^ 2) # pearson chisq 
G.stat
phi = G.stat / m2_fit$df.residual
phi
tilde.phi = m2_fit$deviance / m2_fit$df.residual
tilde.phi 

# test over-dispersion (half normal plot)
res = residuals(m2_fit, type = 'pearson')
plot(qnorm((173 + 1: 173 + 0.5)/(2 * 173 + 1.125)), sort(abs(res)), xlab = 'Expected Half-Normal Order Stats', ylab = 'Ordered Abs Pearson Residuals')
abline(a = 0, b = 1)
abline(a = 0, b = sqrt(phi), lty = 2, col = 'red')
```

There is a linear deviation from the reference line in the half normal plot, suggesting that the response variance of the data exceeds the $\mu$ assumed by the model.   
Hence, we can say that there is over-dispersion in the original model.  
The estimate of dispersion parameter: $\hat\phi=$ `r round(phi, 2)`

```{r}
# adjust for over-dispersion
summary(m2_fit, dispersion = phi)
```

After adjusting for over-dispersion, The coefficient values remain the same. However, the standard errors of each variable differ from the unadjusted model. Now, all coefficients are not statistically significant.

# 2
## a
I will fit a Poisson model with log link with area, year, and leangth of the fish as predictors. Area and Year are treated as categorical variables. 
```{r}
# Poisson model with log link
poi_fit <- glm(Intensity ~ factor(Area) + factor(Year) + Length, family = poisson(link = "log"), data = df_para)
summary(poi_fit)
exp(poi_fit$coefficients)
```

The expected number of parasites given Area = 1, Year = 1999, and Length = 0 is `r round(exp(poi_fit$coefficients)[1], 3)`. All else being equal, the expected number of parasites has multiplicative effect of `r round(exp(poi_fit$coefficients)[2], 3)` on $\mu=E(Y)$ when Area = 2; `r round(exp(poi_fit$coefficients)[3], 3)` when Area = 3; `r round(exp(poi_fit$coefficients)[4], 3)` when Area = 4. Similarly, holding other variables constant, the expected number of parasites has multiplicative effect of `r round(exp(poi_fit$coefficients)[5], 3)` on $\mu=E(Y)$ when Year = 2000; `r round(exp(poi_fit$coefficients)[6], 3)` when Year = 2001. Finally, with every unit increase in Length, the expected number of parasites has multiplicative effect of `r round(exp(poi_fit$coefficients)[7], 3)` on $\mu=E(Y)$ with other predictors unchanged.

## b
```{r}
# Goodness of fit
G = sum(residuals(poi_fit, type = "pearson") ^ 2) 
G 
1 - pchisq(G, poi_fit$df.residual)
```

Given the chi-squared goodness of fit statistic and its p-value, we reject the null and conclude that the model does not have a good fit.

## c
```{r}
# check zero-inflation
check_zeroinflation(poi_fit)
```

I will refit the model using the same predictors accounting for the zero-inflation issue. 

```{r}
# fit zero-inflated poisson model
zip_fit <- zeroinfl(Intensity ~ factor(Area) + factor(Year) + Length, data = df_para) # child and camper for poisson, persons for binary
summary(zip_fit)

exp(coef(zip_fit))
```


All of the predictors in both the count and zero-inflation model are statistically significant except Year = 2001 in count model.  

The zero-inflation model tells us that the baseline odds of being fish that are not susceptible to parasites (Intensity = 0) is `r round(exp(coef(zip_fit)["zero_(Intercept)"]), 3)`. Area 2, 3 (versus Area 1), and Year 2001 (versus 1999) increase the odds of being fish that are susceptible to parasites (Intensity $\neq$ 0) by `r round(exp(coef(zip_fit)["zero_factor(Area)2"]), 3)`, `r round(exp(coef(zip_fit)["zero_factor(Area)3"]), 3)`, `r round(exp(coef(zip_fit)["zero_factor(Year)2001"]), 3)` accordingly.  
The odds is decreased by one unit increase in Length by `r round(exp(coef(zip_fit)["zero_Length"]), 3)`, Area 4 by `r round(exp(coef(zip_fit)["zero_factor(Area)4"]), 3)`, and Year 2000 by `r round(exp(coef(zip_fit)["zero_factor(Year)2000"]), 3)`.  

The count model indicates that the baseline number of parasite is `r round(exp(coef(zip_fit)["count_(Intercept)"]), 3)` among fish with more than one parasites. Area 2, 3, 4 (versus Area 1), and 2000 (versus 1999) increase the number of parasites by `r round(exp(coef(zip_fit)["count_factor(Area)2"]), 3)`, `r round(exp(coef(zip_fit)["count_factor(Area)3"]), 3)`, `r round(exp(coef(zip_fit)["count_factor(Area)4"]), 3)`, `r round(exp(coef(zip_fit)["count_factor(Year)2000"]), 3)` times accordingly. Year 2001 and one unit increase in Length decrease Intensity by `r round(exp(coef(zip_fit)["count_factor(Year)2001"]), 3)` and `r round(exp(coef(zip_fit)["count_Length"]), 3)` times (Year 2001 is insignificant).

